Innovation and elaboration on the avian tree of life - supplementary materials

### 2022-11-11 Manuscript/main_flipped_narrative.tex
Update main_flipped_narrative.tex

Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.
However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general, BurinWhales}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
At the macroevolutionary scale, evolution along a common major axis has been termed elaboration whereas evolution away from the major axis is termed innovation \cite{endler2005animal}. 
The extent to which species and clades elaborate or innovate is poorly understood.
Of particular importance is the potential contribution of reorientations of phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to megaevolutionary scale.
Here we apply new multi-trait methods that model the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
Our analyses show that elaboration dominates species-level phenotypic divergence, consistent with micro-evolutionary theory.
However, we also find that innovation is relatively common at the clade-level suggesting multiple routes to phenotypic innovation throughout avian evolutionary history.
These results show that although phenotypic elaboration and innovation occur across evolutionary scales, elaboration seems to dominate at macroevolutionary scales (i.e. species diverging within clades) and innovation is more prevalent at the megaevolutionary scale (nested patterns of clades diverging within clades).
This suggests that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.
%TG: bleuargh...

% TODO in abstract, add the "paradox" bit: However, if most organisms evolve incrementally along stable lines of least resistance then how can we explain the staggering diversity of species morphologies across the tree of life?

@@

% paragraph 1:  introducing the major axis of variation + novelty and innovation
Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive,Rhoda2022}.
Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
Numerous studies show that phenotypic divergence is biased along the genetic line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales \cite{marroig2005size, mcglothlin2018adaptive, tsuboi2018}.
Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
However, while the line of least genetic resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
The widespread evidence of major shifts in phenotypes across the tree of life \cite{venditti2011multiple,khabbazian2016fast,cooney2017mega,smaers2021evolution,pagel2022general} suggests a tension in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
Resolving this tension requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.

% paragraph 2:  introducing the Endler view.
Endler \textit{et al}. \cite{endler2005animal}, writing in the specific context of signal evolution, describe two routes to phenotypic novelty that we suggest can help to illuminate this challenge.
Specifically, Endler \textit{et al}. refer to evolution of new phenotypes along the same major axis of phenotypic variation as \textit{elaboration}.
This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, would be expected to be a major driver of phenotypic evolution.
In contrast, they refer to evolution away from the major axis of phenotypic variation, either in random directions or along limited axes, as \textit{innovation} \cite{endler2005animal}.
Relative to elaboration, innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
These core concepts can now be defined using robust multivariate statistics about trait covariance applied at multpiple phylogentic scales and provide a powerful theoretical framework with which to study macro- to mega-evolutionary divergence and in particular to determine the relative importance of:
i) idiosyncratic species-level innovation against clade-level innovation (i.e. reorientation of the directions of phenotypic variation),
and ii) evolution along versus away from major axes of phenotypic variation.
The contributions of these multivariate, evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales. 
Here we estimate for the first time, using novel statistical methods, the relative contribution of elaboration and innovation to the remarkable global radiation and diversity of birds beaks.

% paragraph 3: how we measure this (glmm) on the birds data + nestedness
\subsection{Modelling nested trait covariances}
Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
1) large, multivariate datasets to allow exploration of trait covariances at different scales,
2) reliable and efficient computational methods to estimate the major axes of phenotypic variation,
and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric, morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
To meet the second two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the species-level and clade-level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
% start: oldversion
% To provide a baseline from which to calculate elaboration and innovation (i.e. the major axis of phenotypic variation) we use phylogenetic generalised linear mixed models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
% Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components.
% From these models we can calculate and compare major axes of phenotypic variation as the leading eigenvector of the variance-covariance matrix of the trait space.
% end: oldversion
% start: newversion
In order to estimate the major axis of phenotypic variation we fit pGLMM models with beak shape as an eight-dimensional response variable (\cite{MCMCglmm}).
From the posterior distribution of the variance-covariance matrix at any scale, the major axis can be identified from the leading eigenvector of the variance-covariance matrix.
We can estimate levels of elaboration and innovation based on that major axis of phenotypic variation (see Methods).
% end: newversion
We regard the major axis of phenotypic variation as a deep-time analogue of the microevolutionary \textbf{G} matrix \cite{robinson2013quantifying} and posit that it represents the line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
The major axis of phenotypic variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any clade).
To assess the roles of elaboration and innovation across scales we fit our pGLMMs as nested models in which we define variance-covariance matrices, and therefore major axes of phenotypic variation, for
i) the entire phylogeny of all species included in the data (hereafter the global phylogenetic level),
ii) all species within each of nine clades mapping approximately to super-orders (the super-order level),
and iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
These nested partitions of multivariate trait space provide the basis for all subsequent quantitative estimates of elaboration and innovation at multiple scales.

\subsection{Elaboration dominates species-level divergence}
% paragraph 4: results for elaboration and innovation for species
We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along lines of least resistance by calculating species-specific measures of innovation and elaboration.
If microevolutionary lines of least resistance are stable then we expect that species fall on a conserved major axis (elaboration) rather than fall off the line (innovation).
To do this we calculated the multilinear algebraic projection of each species phenotype onto three major axes of phenotypic variation (global phylogenetic, super-order, and order).
We found higher values of elaboration than innovation overall across all levels (Fig \ref{Fig:phylogeny}) and globaly more elaboration and innovation at the global phylogeny and the super-orders levels compared to the orders level.
Phylogenetic distributions of innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or global phylogenetic level (Fig. \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
 and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
However, the extent of both elaboration and innovation tends to reduce as we move from global phylogenetic to within-order comparison.
Broadly, and not surprisingly, this trend reflects the fact that species are more similar to the major axis of phenotypic variation of their close relatives than to the major axis of phenotypic variation of all birds.
The global phylogenetic major axis of phenotypic variation aligns closely with the major axis in the trait space and primarily describes variation between short, deep, and wide beaks at one extreme and long, shallow, and narrow beaks at the other.
Our results point to the possibility that within clades the major axes of phenotypic variation follow different directions.

Overall we found that, as expected from microevolutionary theory, elaboration is the dominant mode of divergence at the species level (Fig. \ref{Fig:phylogeny}).
We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp? 
However, we observe various combinations of elaboration and innovation across birds.
Examples of high innovation and/or elaboration are often clustered within clades.
For example, compared to the global phylogenetic major axis of phenotypic variation, the Trochilidae (hummingbirds) in the order Apodiformes, have long narrow beaks and display high levels of elaboration and low innovation.
In contrast, the Bucerotiformes (hornbills), with large and deep beaks display high levels of innovation and low elaboration.
Other clades such as the Apodidae (swifts), in the order Apodiformes, and the distantly-related but convergent Hirundidae (swallows and martins) in the order Passeriformes, have exceptionally wide beaks relative to length, and display high levels of both elaboration and innovation.
The dominance of elaboration when measured at the species-level is consistent with idea of evolution along lines of least evolutionary resistance.
The frequent phylogenetic clustering of species that elaborate, innovate, or both, highlights the likelihood that major axes of phenotypic variation are not stable among clades.
This leads us to ask: if most organisms evolve incrementally along the major axis of phenotypic variation then how can we explain the staggering diversity of species morphologies across the tree of life?

@@

\noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n = 8748 species) coloured by species beak shape innovation and elaboration level.
Blue and yellow colors on the branches of the phylogeny highlight species with relatively low distances to the global centroid of trait space whereas orange to red colours indicate high distance from the centroid.
Innovation and elaboration scores are represented concentrically in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing comparisons for each species at the order-level, then at the super-order level, and finally at the global phylogenetic level.
The boxplots at the bottom of the figure summarises the distribution of the median innovation and elaboration values for all levels (bottom left) showing that elaboration is more common than elaboration;
and the pooled median innovation and elaboration values for all three different levels (bottom right) showing that elaboration and innovation is more common at higher taxonomic levels than at the orders level.
We can observe an increase in the distance of species to the global centroid of trait space at deeper phylogenetic scales with greater distances generally composed of more elaboration than innovation.
A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.

@@

To investigate the tension implied by our observations of species-level elaboration and previously reported deep time jumps in phenotype, we measured the elaboration and innovation of different clades.

@@

Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to micro-evolutionary theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration in deep time (i.e evidence at the clade scale).

@@

\noindent \textbf{Figure \ref{Fig:orthogonality}:}
Orthogonality of each clade's major axis of phenotypic variation compared to their parent or parent's parent clade.
Orthogonality is represented on the horizontal axis scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines representing, respectively, an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
Dots represent the median orthogonality of each clade and the lines their 95\% CI.
No super-order's major axis of phenotypic variation is parallel to the global phylogenetic major axis (red lines and circles) and 12/25 orders are at least half orthogonal to the major axis of their super-order (green lines and circles).
These results are even clearer when comparing them to the global phylogenetic major axis (green dashed lines and open circles) where only 3/25 are less than half orthogonal.
Furthermore we also indicate here the number of species (n) and the standard deviation of their major axes of phenotypic variation orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
For each clade we also measured the posterior probability of each clade's orientation being different from their parent's clade or the global phylogenetic major axis relative to their sample size and sd.
The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
Only two out of six super-orders; 6/25 orders relative to their super-order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
This is the result of the variation in sample size (n) and standard variation (sd) among clades. 
A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.
%TG: TODO: very easy low hanging fruit (maybe?), but not for this paper is to plot orthogonality through time.

@@

Evolutionary innovation can arise in any direction in trait space.
Although the global phylogenetic major axis of phenotypic variation aligns closely with the first dimension of the raw shapespace (70.96\% of the global phylogenetic major axis of phenotypic variation is aligned on PC1; Fig. \ref{Fig:ellipses}), no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), are aligned mainly with PC1 (Fig. \ref{Fig:ellipses}-bar plots); innovation dominates at this scale.
This pattern of orthogonality also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven out of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes major axis of phenotypic variation (Fig. S\ref{fig_ellipses_passeriformes}).
In addition to the lack of alignment of clade major axes of phenotypic variation with the phylogenetic major axis, we also found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis; e.g. in Galliformes) or higher-dimensional than expected (there is no clear dominant major axis; e.g. in Podicipediformes) from the global phylogenetic major axis of phenotypic variation.
For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the global phylogeny but uniformity in their beak shape components within the clade.
In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
These observations illustrate multiple routes to innovation and imply that within clades species can evolve in many directions (akin to Endler \textit{et al}.'s \cite{endler2005animal} random innovation) or follow a single direction (akin to Endler \textit{et al}.'s \cite{endler2005animal} specific innovation) and that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.

@@

The disconnect between observations of reorientation of trait space at the clade level and a dominance of elaboration at the species level could be viewed as a paradox.
We formalise this paradox by testing whether elaboration is more common at higher taxonomic levels than at the species level.
We compared the area under the curve of the scaled density of elaboration and innovation at both the species level and the clade level.
We then measured the difference between both areas and found that there was a significant difference in innovation among species and clades but no clear difference in elaboration (Fig. \ref{Fig:relative_EI}).
This difference confirms our observation that while innovation is common and perhaps even dominant at the clade level, elaboration dominates at the species level. So, although species evolve preferentially along a shared major axis of phenotypic variance, this major axis changes frequently among clades.
However, we suggest that rather than a paradox the reorientation of trait space can instead arise as a result of small innovations that arise in a common direction among closely-related species \cite{hansen1997stabilizing}.
This idea is similar in concept to models of multiple adaptive peak models where species evolve in a heterogenous adaptive landscape and share similar responses to selection pressures within adaptive zones.
%GT: Tip-toeing into OU territory and needs citations but I think this is a fair suggestion. %TG: fair point!
This implies that gradual directional evolution \cite{pagel2022general}, rather than exceptional megaevolutionary jumps \cite{cooney2017mega,venditti2011multiple}, may be sufficient to explain diversity in avian beak morphology.

@@

\noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among clades (illustrated in figure \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
a) the elaboration (blue) for each species in each 4000 posterior samples compared to the elaboration for each of the 35 clades (yellow).
The scores for each clade are scaled by the maximum score within the clade (corresponding to the scaled ellipses in figure \ref{Fig:ellipses}) and the scores for species are scaled by the maximum species elaboration score.
This allows us to compare elaboration in species and clades despite their different sample sizes.
We use the Bhattacharryya coefficient (BC) to quantify the overlap of the area under the curve for species and clades. 
For elaboration the overlap is BC = 0.217, indicating no distinguishable dissimilarity in the amount of elaboration across the posterior samples between species or clades. 
b) the same as a) but using innovation score. 
For innovation the overlap BC = 0.039, which indicates a clear difference in the amount of innovation across the posterior samples between species and clades.

@@

\textbf{Phylogenetic data.}
We used a random subsample of 1000 trees from the avian tree distribution of \cite{jetz2012global} for the analyses and the one randomly selected tree for the figures (Fig. \ref{Fig:phylogeny}).
We pruned these trees to contain only the 8748 species present in our beak shapespace.
Phylogenetic related data was used in R using the \texttt{ape} package \cite{ape}.

>>>>

Patterns of biological diversity across the tree of life are the result of millions of years of evolutionary history and are shaped by natural selection.
A long-standing proposal is that most morphological diversity at a macroevolutionary scale (among species) arises along a microevolutionary ``line of least resistance''.
However, we frequently observe major shifts in phenotypes among lineages \cite{venditti2011multiple, pagel2022general, BurinWhales}, suggesting that species or clades may also evolve away from the line of least resistance \cite{renaud2006conserved}.
At the macroevolutionary scale, evolution along a common major axis has been termed elaboration whereas evolution away from the major axis is termed innovation \cite{endler2005animal}. 
The extent to which species and clades elaborate or innovate is poorly understood.
Of particular importance is the potential contribution of reorientations of phenotypic trait space (innovation) among clades to contribute to the origins and maintenance of biodiversity across scales from the macro- to mega-evolutionary scale.
Here we apply new multi-trait methods that model the nested covariance structure of traits to evaluate the magnitude and distribution of elaboration and innovation in the evolution of bird beaks.
Our analyses show that elaboration dominates species-level phenotypic divergence, consistent with microevolutionary theory.
However, we also find that innovation is relatively common at the clade-level suggesting multiple routes to phenotypic innovation throughout avian evolutionary history.
These results show that although phenotypic elaboration and innovation occur across evolutionary scales, elaboration seems to dominate at macroevolutionary scales (i.e. species diverging within clades) and innovation is more prevalent at the megaevolutionary scale (nested patterns of clades diverging within clades).
This suggests that both innovation and elaboration are ubiquitous routes for phenotypic change at any scale and that macroevolutionary axes of multivariate evolution are frequently reoriented throughout the history of life, opening up new avenues for evolution to explore.

%TG: cite Anjali somewhere

@@

% paragraph 1:  introducing the major axis of variation + novelty and innovation
Evolutionary theory predicts that over short, microevolutionary, timescales phenotypic evolution will tend to follow evolutionary lines of genetic least resistance \cite{schluter1996adaptive,Rhoda2022}.
Over longer timescales, the evolutionary line of least resistance is analogous to the major axis of phenotypic variation \cite{marroig2005size,fasanelli2022allometry}.
Numerous studies show that phenotypic divergence is biased along the evolutionary line of least resistance, and therefore reflects the major axis of phenotypic variation, over 1-2 million year timescales \cite{marroig2005size, mcglothlin2018adaptive, tsuboi2018}.
Recent evidence suggests that this stability can even extend over macroevolutionary timescales spanning tens of millions of years \cite{mcglothlin2018adaptive}.
However, while the evolutionary line of genetic least resistance explicitly reflects genetic constraints, the macroevolutionary major axis of phenotypic variation is an emergent property of genomic and developmental constraints interacting with selection on a moving adaptive landscape \cite{jones2004evolution}.
The widespread evidence of major shifts in phenotypes across the tree of life \cite{venditti2011multiple,khabbazian2016fast,cooney2017mega,smaers2021evolution,pagel2022general} suggests a conflict in the direction of phenotypic evolution across scales where macroevolutionary divergence away from the major axis of phenotypic variation results in megaevolutionary divergence leading to distinct phenotypes and/or directions of evolution.
Resolving this conflict requires an understanding of the balance between deep-time reorientation of phenotypic major axes of evolution among clades and the directions of evolution of species within clades.

% paragraph 2:  introducing the Endler view.
Endler \textit{et al}. \cite{endler2005animal}, writing in the specific context of signal evolution, describe two routes to phenotypic novelty that we suggest can help to illuminate this challenge.
Specifically, Endler \textit{et al}. refer to evolution of new phenotypes along the same major axis of phenotypic variation as \textit{elaboration}.
This is akin to phenotypic divergence among species following lines of least resistance, and, by definition, would be expected to be a major driver of phenotypic evolution.
In contrast, they refer to evolution away from the major axis of phenotypic variation, either in random directions or along limited axes, as \textit{innovation} \cite{endler2005animal}.
Relative to elaboration, innovation may be more likely to generate phenotypic novelty, but this need not be synonymous with being a primary driver of phenotypic diversity at broad scales.
These core concepts can now be defined using robust multivariate statistics about trait covariance applied at multiple phylogenetic scales and provide a powerful theoretical framework with which to study macro- to mega-evolutionary divergence and in particular to determine the relative importance of:
i) idiosyncratic species-level innovation against clade-level innovation (i.e. reorientation of the directions of phenotypic variation),
and ii) evolution along versus away from major axes of phenotypic variation.
The contributions of these multivariate, evolutionary pathways are poorly-known despite their importance for understanding the origins of diversity from macro- to mega-evolutionary scales. 
Here we estimate for the first time, using novel statistical methods, the relative contribution of elaboration and innovation to the remarkable global radiation and diversity of birds beaks.

% paragraph 3: how we measure this (glmm) on the birds data + nestedness
\subsection{Modelling nested trait covariances}
Addressing the relative contributions of elaboration and innovation to the origins of biodiversity in deep-time requires
1) large, multivariate datasets to allow exploration of trait covariances at different scales,
2) reliable and efficient computational methods to estimate the major axes of phenotypic variation,
and 3) a set of mathematical tools that can estimate degrees of elaboration and innovation at any scale.
To meet the first challenge we use an eight-dimensional beak shape morphospace based on a geometric, morphometric dataset of 8748 species of birds (described in \cite{hughes2022global}).
To meet the other two challenges we introduce a novel analytical pipeline for measuring elaboration and innovation at the species-level and clade-level (Fig.S\ref{Fig:cheat_sheet} and S\ref{Fig:mcmcmcglmm}).
% start: oldversion
% To provide a baseline from which to calculate elaboration and innovation (i.e. the major axis of phenotypic variation) we use phylogenetic generalised linear mixed models (pGLMM) with beak shape as an eight-dimensional response variable (see Methods; \cite{MCMCglmm}).
% Such models allow us to measure variance-covariance matrices partitioned into phylogenetic and non-phylogenetic components.
% From these models we can calculate and compare major axes of phenotypic variation as the leading eigenvector of the variance-covariance matrix of the trait space.
% end: oldversion
% start: newversion
To estimate the major axis of phenotypic variation we fit Bayesian phylogenetic generalised linear mixed model (pGLMM) models with beak shape as an eight-dimensional response variable (\cite{MCMCglmm}).
From the posterior distribution of the variance-covariance matrix at any scale, the major axis can be identified from the leading eigenvector of the variance-covariance matrix.
We can estimate levels of elaboration and innovation based on that major axis of phenotypic variation (see Methods).
% end: newversion
We regard the major axis of phenotypic variation as a deep-time analogue of the microevolutionary \textbf{G} matrix \cite{robinson2013quantifying} and posit that it represents the line of least evolutionary resistance, capturing the effects of historical contingency on multivariate evolution.
The major axis of phenotypic variation can be defined at any phylogenetic scale (i.e. for all species in the phylogeny, or for all species within any clade).
To assess the roles of elaboration and innovation across scales we fit our pGLMMs as nested models in which we define variance-covariance matrices, and therefore major axes of phenotypic variation, for
i) the entire phylogeny of all species included in the data (hereafter the global phylogenetic level),
ii) all species within each of nine clades mapping approximately to super-orders (the super-order level),
and iii) all species within each of 27 clades mapping to taxonomic orders (the order level).
These nested partitions of multivariate trait space provide the basis for all subsequent quantitative estimates of elaboration and innovation at multiple scales.

\subsection{Elaboration dominates species-level divergence}
% paragraph 4: results for elaboration and innovation for species
We tested the expectation, derived from adaptive radiation theory, that species divergence is biased along evolutionary lines of least resistance by calculating species-specific measures of innovation and elaboration.
If microevolutionary lines of least resistance are stable then we expect that species fall on a conserved major axis (elaboration) rather than off the line (innovation).
To do this we calculated the multilinear algebraic projection of each species phenotype onto three major axes of phenotypic variation (global phylogenetic, super-order, and order).
We found higher values of elaboration than innovation overall across all levels (Fig \ref{Fig:phylogeny}) and globally more elaboration and innovation at the global phylogenetic and super-order levels compared to the order level.
Phylogenetic distributions of innovation and elaboration broadly hold whether comparisons are made at the order, super-order, or global phylogenetic level (Fig. \ref{Fig:phylogeny})%{#fig_correlations}),%TODO: maybe add to supp?
 and in further nested analysis of the order Passeriformes (Fig. S\ref{fig_ellipses_passeriformes} and S\ref{fig_phylogeny_passeriformes}).
However, the extent of both elaboration and innovation tends to reduce as we move from global phylogenetic to within-order comparisons.
Broadly, and not surprisingly, this trend reflects the fact that species are more similar to the major axis of phenotypic variation of their close relatives than to the major axis of phenotypic variation of all birds.
The global phylogenetic major axis of phenotypic variation aligns closely with the major axis in the trait space and primarily describes variation between short, deep, and wide beaks at one extreme, and long, shallow, and narrow beaks at the other.
Our results point to the possibility that within clades the major axes of phenotypic variation follow different directions.

Overall we found that, as expected from microevolutionary theory, elaboration is the dominant mode of divergence at the species level (Fig. \ref{Fig:phylogeny}).
We find no consistent evidence that innovation is either positively correlated with, or trades-off with, elaboration. %(Fig. S{#fig_correlations}). %TODO: maybe add to supp? 
However, we observe various combinations of elaboration and innovation across birds.
Examples of high innovation and/or elaboration are often clustered within clades.
For example, compared to the global phylogenetic major axis of phenotypic variation, the Trochilidae (hummingbirds) in the order Apodiformes, have long narrow beaks and display high levels of elaboration and low innovation.
In contrast, the Bucerotiformes (hornbills), with large and deep beaks display high levels of innovation and low elaboration.
Other clades such as the Apodidae (swifts), in the order Apodiformes, and the distantly-related but convergent Hirundinidae (swallows and martins) in the order Passeriformes, have exceptionally wide beaks relative to length, and display high levels of both elaboration and innovation.
The dominance of elaboration when measured at the species-level is consistent with idea of evolution along lines of least evolutionary resistance.
The frequent phylogenetic clustering of species that elaborate, innovate, or both, highlights the likelihood that major axes of phenotypic variation are not stable among clades.
This leads us to ask; if most organisms evolve incrementally along the major axis of phenotypic variation then how can we explain the staggering diversity of species morphologies across the tree of life?

@@

\noindent \textbf{Figure \ref{Fig:phylogeny}:} The avian phylogeny (n = 8748 species) coloured by species beak shape innovation and elaboration level.
Blue and yellow colours on the branches of the phylogeny highlight species with relatively low distances to the global centroid of trait space whereas orange to red colours indicate high distance from the centroid.
Innovation and elaboration scores are represented concentrically in a nested fashion with the most external pairs of circles (innovation and elaboration bands) representing comparisons for each species at the order level, then at the super-order level, and finally at the global phylogenetic level.
The boxplots at the bottom of the figure summarise the distribution of the median innovation and elaboration values for all levels (bottom left), showing that elaboration is more common than elaboration,
and the pooled median innovation and elaboration values for all three different levels (bottom right), showing that elaboration and innovation is more common at higher taxonomic levels than at the order level.
We can observe an increase in the distance of species to the global centroid of trait space at deeper phylogenetic scales with greater distances generally composed of more elaboration than innovation.
A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_phylogeny_passeriformes}.

@@

To investigate the conflict implied by our observations of species-level elaboration and previously reported deep-time jumps in phenotype, we measured the elaboration and innovation of different clades.

@@

Remarkably, more than half of the assessed clades (4/8 super-orders and 15/27 orders) displayed higher median innovation than median elaboration scores implying that, contrary to microevolutionary theory \cite{schluter1996adaptive,marroig2005size}, innovation is a more common generator of morphological diversity than elaboration in deep-time (i.e evidence at the clade scale).

@@

\noindent \textbf{Figure \ref{Fig:orthogonality}:}
Orthogonality of each clade's major axis of phenotypic variation compared to their parent or parent's parent clade.
Orthogonality is represented on the horizontal axis and scales from 0 (modulo of $0^\circ$) to 1 (modulo of $90^\circ$) with the background grey and dashed grey lines representing, respectively, an orthogonality of 0.5 (modulo of $45^\circ$) and 0.75 (modulo of $67.5^\circ$).
Dots represent the median orthogonality of each clade and the lines their 95\% CI.
No super-order's major axis of phenotypic variation is parallel to the global phylogenetic major axis (red lines and circles) and 12/25 orders are at least half orthogonal to the major axis of their super-order (green lines and circles).
These results are even clearer when compared to the global phylogenetic major axis (green dashed lines and open circles) where only 3/25 are less than half orthogonal.
We also indicate the number of species (n) and the standard deviation (sd) of their major axes of phenotypic variation orientation over the 4000 variance-covariance posteriors (sd; expressed in degrees).
For each clade we also measured the posterior probability of each clade's orientation being different from their parent's clade or the global phylogenetic major axis relative to their sample size and sd.
The stars represent the posterior probability of the clade's orientation being different from the comparison clade. (*** = pp $> 0.99$; ** = pp $>0.95$; * = pp $> 0.9$; . = pp $> 0.8$).
Only two out of six super-orders; 6/25 orders relative to their super-order and 9/25 orders relative to the phylogeny have a pp $> 0.99$.
This is the result of the variation in sample size (n) and standard variation (sd) among clades. 
A companion plot showing further nested structure within the Passeriformes is shown in Fig. S\ref{fig_orthogonality_passeriformes}.
%TG: TODO: very easy low hanging fruit (maybe?), but not for this paper is to plot orthogonality through time.

@@

Evolutionary innovation can arise in any direction in trait space.
Although the global phylogenetic major axis of phenotypic variation aligns closely with the first dimension of the raw shapespace (70.96\% of the global phylogenetic major axis of phenotypic variation is aligned with PC1; Fig. \ref{Fig:ellipses}), no super-order, and only seven of the 27 orders (Procellariiformes, Pelecaniformes, Charadriiformes, Coraciiformes, Psittaciformes and Passeriformes), are aligned mainly with PC1 (Fig. \ref{Fig:ellipses}-bar plots); innovation dominates at this scale.
This pattern of orthogonality also holds for sub-orders and families within the Passeriformes where only half of the sub-orders (Meliphagoidea, Corvides and Passerida) and seven of the 23 families (Eurylaimides, Meliphagoidea, Petroicidea, Fringilidae, Aegithaloidea, Pyconotiae, Nectariniidae) align with the Passeriformes major axis of phenotypic variation (Fig. S\ref{fig_ellipses_passeriformes}).
In addition to the lack of alignment of clade major axes of phenotypic variation with the phylogenetic major axis, we also found that within clades, phenotypic variation is either highly constrained (varying almost entirely along a single axis; e.g. in Galliformes) or higher-dimensional than expected (there is no clear dominant major axis; e.g. in Podicipediformes) from the global phylogenetic major axis of phenotypic variation.
For example, Accipitriformes (hawks and allies) have major axes of phenotypic variation that mainly align with the second dimension of the shapespace suggesting distinct directions of evolution for the clade relative to the global phylogenetic level but uniformity in their beak shape within the clade.
In contrast, the Podicipediformes (grebes) are highly variable across all dimensions suggesting that all components of the shapespace are necessary to describe their beaks.
These observations illustrate multiple routes to innovation and imply that within clades species can evolve in many directions (akin to Endler \textit{et al}.'s \cite{endler2005animal} random innovation) or follow a single direction (akin to Endler \textit{et al}.'s \cite{endler2005animal} specific innovation) and that reorientations of trait space can arise in any direction, in any lineage, and at any time throughout avian history.

@@

The disconnect between observations of reorientation of trait space at the clade level and a dominance of elaboration at the species level could be viewed as a paradox.
We formalise this paradox by testing whether elaboration is more common at higher taxonomic levels than at the species level.
We compared the area under the curve of the scaled density of elaboration and innovation at both the species level and the clade level.
We then measured the difference between both areas and found that there was a significant difference in innovation among species and clades but no clear difference in elaboration (Fig. \ref{Fig:relative_EI}).
This difference confirms our observation that while innovation is common and perhaps even dominant at the clade level, elaboration dominates at the species level. So, although species evolve preferentially along a shared major axis of phenotypic variance, this major axis changes frequently among clades.
However, we suggest that rather than a paradox the reorientation of trait space can instead arise as a result of small innovations that arise in a common direction among closely-related species.
This idea is similar in concept to multiple adaptive peak models where species evolve in a heterogeneous adaptive landscape and share similar responses to selection pressures within adaptive zones \cite{hansen1997stabilizing}.
%GT: Tip-toeing into OU territory and needs citations but I think this is a fair suggestion. %TG: fair point!
This implies that gradual directional evolution \cite{pagel2022general}, rather than exceptional megaevolutionary jumps \cite{cooney2017mega,venditti2011multiple}, may be sufficient to explain diversity in avian beak morphology.

@@

\noindent \textbf{Figure \ref{Fig:relative_EI}:} Comparisons of relative elaboration and innovation among clades (illustrated in figure \ref{Fig:ellipses}) and among species (figure \ref{Fig:phylogeny}).
a) the elaboration (blue) for each species in each 4000 posterior samples compared to the elaboration for each of the 35 clades (yellow).
The scores for each clade are scaled by the maximum score within the clade (corresponding to the scaled ellipses in figure \ref{Fig:ellipses}) and the scores for species are scaled by the maximum species elaboration score.
This allows us to compare elaboration in species and clades despite their different sample sizes.
We use the Bhattacharryya coefficient (BC) to quantify the overlap of the area under the curve for species and clades. 
For elaboration the overlap is BC = 0.217, indicating no distinguishable dissimilarity in the amount of elaboration across the posterior samples between species or clades. 
b) the same as a) but using innovation score. 
For innovation the overlap has BC = 0.039, which indicates a clear difference in the amount of innovation across the posterior samples between species and clades.

@@

\textbf{Phylogenetic data.}
We used a random subsample of 1000 trees from the avian tree distribution of \cite{jetz2012global} for the analyses and the one randomly selected tree for the figures (Fig. \ref{Fig:phylogeny}).
We pruned these trees to contain only the 8748 species present in our beak shapespace.
